Compact orbitals enable low-cost linear-scaling \emph{ab initio} molecular dynamics for weakly-interacting systems

### 2018-05-28 JCP/SI.tex

\title{
Supporting Information: \\
Compact orbitals enable low-cost linear-scaling \emph{ab initio} molecular dynamics for weakly-interacting systems
}

@@

As described in Refs.~\citenum{almo-ls}, ALMO DFT is designed to be computationally efficient for large systems that can be logically partitioned into small fragments---molecules, atoms, ions---not connected to each other by strong covalent bonds. ALMO DFT differs from the conventional DFT only by constraints imposed on electrons: all electrons and AOs in the system are assigned to fragments and each electron is allowed to delocalize only over its own fragment and all neighboring fragments within a pre-defined cutoff radius $R_c$ (Figure~\ref{fig:blocks}A). This restriction imposes the blocked pattern on electron's molecular orbital and the molecular orbital coefficient matrix $\mathbf{T}$ (Figure~\ref{fig:blocks}B).

@@

The algorithms for the force evaluation rely on the following conventions. Functions with names starting with \texttt{ALMO} are written out explicitly. Functions with names starting with \texttt{DFT} are conventional and are not shown. There are two functions that refer to the direct inversion in the iterative subspace (DIIS) method~\cite{ZZZ}. \texttt{DIIS.push} updates the DIIS matrix and enables \texttt{DIIS.extrapolate} to generate a new variable that, hopefully, minimizes the error. These functions are not written out explicitly. Functions with names starting with \texttt{HISTORY} store and retrieve the contravariant-covariant density matrices from recent AIMD steps. These matrices are necessary to generate a good initial guess for the electronic descriptors in the two self-consistent field (SCF) procedures. The atomic orbital (AO) overlap matrix $\mathbf{S}$ remains constant within one AIMD step and, therefore, is conveniently treated as a global variable visible in all procedures.

@@

The efficiency of both SCF stages is greatly improved by the availability of a good initial guess for the localized electronic descriptors. The initial guess is generated using an algorithm (Figure~\ref{fig:guess}) inspired by the always stable predictor-corrector method of Kolafa~\cite{Kolafa2004}.

@@

	\State $\mathbf{return}$ $\mathbf{T}_{R_c}$
	\EndProcedure
   \end{algorithmic}
\end{algorithm}
\caption{\label{fig:guess} Initial guess for the localized electronic descriptors. The algorithm is inspired by always stable predictor-corrector method of Kolafa~\cite{Kolafa2004}.}
\end{figure}

@@

>>>>

\title{
Supplementary Material: \\
Compact orbitals enable low-cost linear-scaling \emph{ab initio} molecular dynamics for weakly-interacting systems
}

@@

As described in Refs.~\citenum{a:almo-ls}, ALMO DFT is designed to be computationally efficient for large systems that can be logically partitioned into small fragments---molecules, atoms, ions---not connected to each other by strong covalent bonds. ALMO DFT differs from the conventional DFT only by constraints imposed on electrons: all electrons and AOs in the system are assigned to fragments and each electron is allowed to delocalize only over its own fragment and all neighboring fragments within a pre-defined cutoff radius $R_c$ (Figure~\ref{fig:blocks}A). This restriction imposes the blocked pattern on electron's molecular orbital and the molecular orbital coefficient matrix $\mathbf{T}$ (Figure~\ref{fig:blocks}B).

@@

The algorithms for the force evaluation rely on the following conventions. Functions with names starting with \texttt{ALMO} are written out explicitly. Functions with names starting with \texttt{DFT} are conventional and are not shown. There are two functions that refer to the direct inversion in the iterative subspace (DIIS) method. \texttt{DIIS.push} updates the DIIS matrix and enables \texttt{DIIS.extrapolate} to generate a new variable that, hopefully, minimizes the error. These functions are not written out explicitly. Functions with names starting with \texttt{HISTORY} store and retrieve the contravariant-covariant density matrices from recent AIMD steps. These matrices are necessary to generate a good initial guess for the electronic descriptors in the two self-consistent field (SCF) procedures. The atomic orbital (AO) overlap matrix $\mathbf{S}$ remains constant within one AIMD step and, therefore, is conveniently treated as a global variable visible in all procedures.

@@

The efficiency of both SCF stages is greatly improved by the availability of a good initial guess for the localized electronic descriptors. The initial guess is generated using an algorithm (Figure~\ref{fig:guess}) inspired by the always stable predictor-corrector method of Kolafa~\cite{Kolafa2003}.

@@

	\State $\mathbf{return}$ $\mathbf{T}_{R_c}$
	\EndProcedure
   \end{algorithmic}
\end{algorithm}
\caption{\label{fig:guess} Initial guess for the localized electronic descriptors. The algorithm is inspired by always stable predictor-corrector method of Kolafa~\cite{Kolafa2003}.}
\end{figure}

@@

\bibliography{SI}

### 2018-05-28 JCP/manuscript.tex

The new AIMD method utilizes a recently developed LS DFT~\cite{a:almo-ls} based on absolutely localized molecular orbitals (\mbox{ALMOs}) -- compact orbitals first described in Ref.~\citenum{a:stoll}. 
Unlike delocalized KS orbitals, each \mbox{ALMO} has its own \emph{localization center} and a predefined \emph{localization radius} $R_{c}$ that typically includes nearby atoms or molecules~\cite{a:stoll,a:almo-ls}. 
In the current implementation, a localization center is defined as a set of all Gaussian atomic orbitals of one molecule. 
However, the approach can use other local and nonlocal basis sets~\cite{a:galli_loc, Lin2012}. 
The key feature of ALMO DFT is that its one-electron wavefunctions are constructed in a two-stage self-consistent-field (SCF) procedure~\cite{a:almo-ls} to circumvent the problem of the sluggish variational optimization emphasized above. 
In the first stage, ALMOs are constrained to their localization centers~\cite{a:khal} whereas, in the second stage, ALMOs are relaxed to allow delocalization onto the neighbor molecules within their localization radius $R_{c}$. 
To achieve a robust optimization in the problematic second stage, it is important to keep the delocalization component of the trial wavefunction orthogonal to the fixed orbitals obtained in the first stage. 
For mathematical details, see the ALMO SCF method in Ref.~\citenum{a:almo-ls}. ALMO DFT agorithms are presented in the Supporting Information.

@@

ALMO AIMD was implemented in CP2K, an open source materials modeling package~\cite{www:cp2k}. 
Accuracy and efficiency of ALMO AIMD was tested using liquid water as an example. 
This system is challenging because intermolecular electron delocalization is a critical component of hydrogen bonding and must be described correctly to reproduce static and dynamical properties of liquid water. 
A periodic cell containing 125 molecules was simulated for $\SI{30}{\ps}$ at $T=\SI{298}{\K}$ and a constant density of $\SI{1.01}{\g\cdot\cm^{-3}}$. 
The initial configurations of the water boxes were created by equilibrating the system for 20~ps using fully converged orbitals. 
Ricci-Ciccotti algorithm~\cite{Ricci2003} was used to integrate the Langevin equation (see Supporting Information). We found that $\gamma = \SI{e-3}{\per\fs}$ is large enough to thermostat the system efficiently and small enough not to significantly affect dynamical properties of liquid water.
%
In the dual Gaussian and plane-wave scheme implemented in CP2K~\cite{a:quickstep}, the TZV2P basis set was used to represent molecular orbitals, and a plane-wave cutoff of $\SI{320}{Ry}$ used to represent electron density. 
The XC energy was approximated using the dispersion-corrected PBE functional~\cite{a:pbe,Grimme2010}. 
Separable norm-conserving pseudopotentials were used~\cite{a:hgh} and the Brillouin zone was sampled at the $\Gamma$-point. 
%Although not a required component of ALMO AIMD, 
The predictor of the Kolafa scheme~\cite{Kolafa2003} was adopted to localized orbitals~\cite{a:2ndcpmd} to generate a highly accurate initial ALMOs in both SCF stages, which can be brought close to the ground state with just a few SCF steps of the robust two-stage optimization procedure. 
%RZK: The time-reversibility of ASPC is not a critical reqirement of the proposed method.

@@

Even with $\epsilon_{\text{SCF}} = 10^{-2}$, the simulation is stable with the correct average temperature and perfect Maxwell-Boltzmann distribution (Figure~\ref{fig:dynproperties}b). $\Delta$ was initially estimated at $\SI{2e-5}{\per\fs}$ using Eq.~(\ref{eq:delta}) and then refined heuristically to $\SI{6e-5}{\per\fs}$. 
We found that it is easier to optimize $\Delta$ when $\gamma$ is set to zero because of reduced noise in the trial runs. 
%RZK addition for a longer article: In theory, $\Delta$ should be independent on system size. However, we found that $\Delta$ dependents on system size due to finite size effects, which decay for increasing system size.
%
Analysis of $\delta \vec{f_{i}}(t)$ shows that the error indeed resembles Gaussian white noise. The mean of the error is zero (black line in Figure~\ref{fig:randomforce}a). Its ACF decays rapidly (Figure~\ref{fig:randomforce}c) so that the errors can be considered uncorrelated on time scale of $\SI{50}{\fs}$. Thus the main assumption behind our approach to ALMO AIMD is justified for liquid water. We established that the main source of error in forces for this system is the loose convergence criterion and not the finite $R_c$: fully converged ALMO SCF calculations remove the oscillating component of $\delta f$ (Figure~\ref{fig:randomforce}b). We also verified that the ALMO forces converge to the reference forces in the limit $R_{c} \rightarrow \infty$ 
%(Figure~\ref{fig:forcecomp} in the Supplemental Material).
(Figure~\ref{SI-fig:forcecomp} in the Supporting Information).

@@

%Convergence of approximate forces to the reference forces with the electron localization radius.

@@

\textbf{Acknowledgements.} The authors are grateful to Thomas K\"uhne for insightful discussions. The research was funded by the Natural Sciences and Engineering Research Council of Canada through the Discovery Grant (RGPIN-2016-05059). The authors are grateful to Compute Canada and McGill HPC Centre for computer time.

>>>>

The new AIMD method utilizes a recently developed LS DFT~\cite{a:almo-ls} based on absolutely localized molecular orbitals (\mbox{ALMOs}) -- compact orbitals first described in Ref.~\citenum{a:stoll}. 
Unlike delocalized KS orbitals, each \mbox{ALMO} has its own \emph{localization center} and a predefined \emph{localization radius} $R_{c}$ that typically includes nearby atoms or molecules~\cite{a:stoll,a:almo-ls}. 
In the current implementation, a localization center is defined as a set of all Gaussian atomic orbitals of one molecule. 
However, the approach can use other local and nonlocal basis sets~\cite{a:galli_loc, Lin2012}. 
The key feature of ALMO DFT is that its one-electron wavefunctions are constructed in a two-stage self-consistent-field (SCF) procedure~\cite{a:almo-ls} to circumvent the problem of the sluggish variational optimization emphasized above. 
In the first stage, ALMOs are constrained to their localization centers~\cite{a:khal} whereas, in the second stage, ALMOs are relaxed to allow delocalization onto the neighbor molecules within their localization radius $R_{c}$. 
To achieve a robust optimization in the problematic second stage, it is important to keep the delocalization component of the trial wavefunction orthogonal to the fixed orbitals obtained in the first stage.

@@

ALMO AIMD was implemented in CP2K, an open source materials modeling package~\cite{www:cp2k}. 
Accuracy and efficiency of ALMO AIMD was tested using liquid water as an example. 
This system is challenging because intermolecular electron delocalization is a critical component of hydrogen bonding and must be described correctly to reproduce static and dynamical properties of liquid water. 
A periodic cell containing 125 molecules was simulated for $\SI{30}{\ps}$ at $T=\SI{298}{\K}$ and a constant density of $\SI{1.01}{\g\cdot\cm^{-3}}$. 
The initial configurations of the water boxes were created by equilibrating the system for 20~ps using fully converged orbitals. 
%
In the dual Gaussian and plane-wave scheme implemented in CP2K~\cite{a:quickstep}, the TZV2P basis set was used to represent molecular orbitals, and a plane-wave cutoff of $\SI{320}{Ry}$ used to represent electron density. 
The XC energy was approximated using the dispersion-corrected PBE functional~\cite{a:pbe,Grimme2010}. 
Separable norm-conserving pseudopotentials were used~\cite{a:hgh} and the Brillouin zone was sampled at the $\Gamma$-point. 
%Although not a required component of ALMO AIMD, 
The predictor of the Kolafa scheme~\cite{Kolafa2003} was adopted to localized orbitals~\cite{a:2ndcpmd} to generate a highly accurate initial ALMOs in both SCF stages, which can be brought close to the ground state with just a few SCF steps of the robust two-stage optimization procedure. 
%RZK: The time-reversibility of ASPC is not a critical reqirement of the proposed method.

@@

Even with $\epsilon_{\text{SCF}} = 10^{-2}$, the simulation is stable with the correct average temperature and perfect Maxwell-Boltzmann distribution (Figure~\ref{fig:dynproperties}b). $\Delta$ was initially estimated at $\SI{2e-5}{\per\fs}$ using Eq.~(\ref{eq:delta}) and then refined heuristically to $\SI{6e-5}{\per\fs}$. 
We found that it is easier to optimize $\Delta$ when $\gamma$ is set to zero because of reduced noise in the trial runs. 
%RZK addition for a longer article: In theory, $\Delta$ should be independent on system size. However, we found that $\Delta$ dependents on system size due to finite size effects, which decay for increasing system size.
%

@@

@@
